# Kaplan Meier

## Kaplan Meier Survival Estimate

The Kaplan Meier survival estimate is how survival probabilities can be calculated while factoring in censored times. The Kaplan Meier survival estimate is used to calculate probability of survival at a given time where $S(t_j)$ is the probability of being alive at time $t_j$, $S(t_{j-1})$ is the probability of being alive at $t_{j-1}$, $n_j$ is the number of patients alive just before $t_j$, $d_j$ is the number of events at $t_j$, and $j$ is the time interval of interest. The equation is $S(t_j)=S(t_{j-1})(1-\frac{d_j}{n_j})$ (@clark03). The equation essentially divides the surviving individuals by the individuals at risk, similar to the previous calculations shown. However, Kaplan Meier curves adjust for right-censored times by dropping observations from the total number of individuals at risk, $n_j$, after their censored time has been reached. This adjustment prevents overestimating the survival probability because it no longer assumes censored individuals are still alive or and at risk (@goel10).

## Kaplan Meier Curve and Descriptive Statistics

The Kaplan Meier curve is a graphical representation of survival analysis. Similar to the survival probabilities discussed previously, the Kaplan Meier curve shows the relationships between time, which is typically plotted on the x-axis, and probability of survival, which is typically on the y-axis. The curve always ranges from 0 to 1 and is right skewed. Useful summary statistics for a Kaplan Meier Survival curve include confidence intervals and the median. A 95% confidence interval for each point can be found by using the formula $S_t ± 1.96 * SE(S_t)$. The median value is typically reported rather than the mean because mean survival time cannot be reported reliably for those who have not experienced the outcome of interest yet. The median can be found by finding the time when probability of survival is equal to 0.5. Thus, the median can only be reported when at least half of the participants experienced the outcome of interest during the study (@rao23).

## Kaplan Meier in R

Luckily, modern software makes these calculations easy and fast, as well as plotting them with confidence intervals and risk tables. A new data set with censored times will be created to demonstrate this process. To do this, a 0 will be recorded for some individuals at times before t = 10. @tbl-censor shows this data set's structure.

```{r}
# Load Packages
library(tidyverse) |> suppressPackageStartupMessages()
library(knitr)
library(survival)
library(ggsurvfit)
library(gt)
```

```{r}
# Create data set
id <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
time <- c(1, 2, 10, 4, 5, 6, 10, 8, 9, 10)
status <- c(1, 1, 0, 0, 1, 0, 0, 1, 1, 0)
censor <- data.frame(id, time, status)
```

```{r}
#| label: tbl-censor
#| tbl-cap: "Example Data Set with Censored Times"

censor %>% gt(caption = "Example Data Set with Censored Times") %>%
  cols_label(id = "ID", time = "Time", status = "Status") %>%
  tab_source_note(source_note = "Table 3") 
```

The `survival` package in `R` has a function called `Surv()` that takes input data and creates a response object recording survival time for each observation. The function takes in the time variable and the status variable. The function takes into account right censoring as well, marking censored times with a $+$ symbol in the object created (@zabor23).

```{r}
# Show the time of event or censored time for each observation
times <- Surv(censor$time, censor$status)
times
```

The `survfit()` function can be used to calculate the Kaplan Meier survival estimate for each time that a new event occurs. The function takes in the response object created by the `Surv()` function in order to drop censored observations from the regression. We use \~ 1 because we are not including predictor variables in the model. The function returns a survival object that can be used to plot the Kaplan Meier curve and calculate summary statistics. The `time` variable in the object created by the `survfit()` function shows the time of each event, and the `surv` variable shows the survival probability for the remaining individuals after each event occurs.

```{r}
# Calculate survival object
s1 <- survfit(times ~ 1, data = censor)
# View the survival time for each time an event occurs
s1$time
# View survival probability for remaining individuals after each event occurs 
round(s1$surv, 2)
```

The `ggsurvfit` package can be used to plot the Kaplan Meier curve for this data using the previously created `s1` object. @fig-survcurv shows the Kaplan Meier curve for predicting the event. The risk table shows the number of individuals at risk at each time point and the number of events that occurred at each time point.

```{r}
#| label: fig-survcurv
#| fig-cap: "Kaplan Meier Survival Curve"

s1 %>% 
  ggsurvfit() +
  labs(
    x = "Time",
    y = "Overall survival probability",
    title = "Kaplan Meier Survival Curve"
  ) + 
  scale_x_continuous(breaks=seq(0,10,by=2)) + 
  scale_y_continuous(breaks=seq(0,1,by=.2)) +
  add_risktable()
```

## Log-Rank Test

One of the most common applications of survival analysis and Kaplan Meier curves is for comparing survival times between two groups. One example of when this might be useful is if a study is comparing survival after two different treatment plans. A data set with two groups will be created to demonstrate this process.

```{r}
# Create new data set
time <- c(8, 7, 10, 8, 5, 3, 4, 10, 6, 1)
status <- c(0, 1, 0, 1, 0, 1, 0, 0, 1, 1)
group <- c(1, 1, 1, 1, 1, 2, 2 ,2, 2, 2)
surv2 <- data.frame(time, status, group)
```

To run analysis, the statistical test called the Log-Rank Test can be used to test the null hypothesis that there is no difference between the survival estimates of two groups at any point in time (@rich10). The test is conducted by comparing the observed number of event with an estimated number of events for each group at each time. To model this, we need to record some values of interest from our existing data. The first, $n1t$, represents the number of people alive and still in the study, or at risk, at time t in group 1. Similarly, $n2t$ is the number at risk in group 2 at time t. $nt$ is the sum of these two values at time t, representing the total number of people at risk in the study at time t. $o1t$ is the number of observed events occurring at time t in group 1, $o2t$ is observed events occurring at time t in group 2, and $ot$ is the total observed events across both groups at time t. To calculate expected number of events, the assumption that the two curves are identical is used. The expected number of events at a time t is for group i calculated by the formula $E_{it} = \frac{N_{it} \times O_{t}}{N_{t}}$, where $N_{it}$ is the observed number of events in group i at time t, $O_{t}$ is the total number of events in all groups at time t, and $N_{t}$ is the total number observations at risk in all groups at time t (@sullivan16). @tbl-logrank shows the creation of all of these values, where $e1t$ is the expected number of events at time t in group 1, $e2t$ is the expected number of events at time t in group 2, and $et$ is the total expected number of events across both groups at time t. Cumulative sums are also calculated for each of these values.

```{r}
#| label: tbl-logrank
#| tbl-cap: "Calculation of Log-Rank Test Statistic"

Time <- c(1, 3, 4, 5, 6, 7, 8, 10)
N1t <- c(5, 5, 5, 5, 4, 4, 3, 1)
N2t <- c(5, 4, 3, 2, 2, 1, 1, 1)
O1t <- c(0, 0, 0, 0, 0, 1, 1, 0)
O2t <- c(1, 1, 0, 0, 1, 0, 0, 0)

log_rank <- data.frame(Time, N1t, N2t, O1t, O2t)

# Calculate Totals
log_rank$Nt <- log_rank$N1t + log_rank$N2t
log_rank$Ot <- log_rank$O1t + log_rank$O2t
log_rank$E1t <- (log_rank$N1t * log_rank$Ot) / log_rank$Nt
log_rank$E2t <- (log_rank$N2t * log_rank$Ot) / log_rank$Nt

round(log_rank, 2) %>% 
  gt() %>% 
  tab_header("Values Needed to Calculate Chi-Square Statistic") %>% 
  cols_label(Time = "Time", 
             N1t = "Number at Risk in Group 1", 
             N2t = "Number at Risk in Group 2", 
             Nt = "Total Number at Risk", 
             O1t = "Observed Events in Group 1", 
             O2t = "Observed Events in Group 2", 
             Ot = "Total Observed Events", 
             E1t = "Expected Events in Group 1", 
             E2t = "Expected Events in Group 2") 
```

To test the null hypothesis that there is no difference between the survival estimates of two groups at any point in time, a test statistics is needed. For the Log-Rank test, a Chi-Square test statistic is used because the data follows a Chi-Square curve rather than a normal distribution. The Chi-Square test statistic is calculated by the formula $X^2 = \sum_{i=1}^k \frac{(\sum{O_{it}} - \sum{E_{it}})^2}{\sum{Vi}}$, where $\sum_{0}^T{O_{it}}$ is the sum of the observed number of events in group i over the entire time interval and $\sum_{0}^T{E_{it}}$ is the sum of the expected number of events in group i over the entire time interval. The difference of the sums of these two values are divided by the sum of the variance. The variance for group 1 at time t is calculated by the formula $V_{1t} = \frac{N_{1t} \times N_{2t} \times O_{t} \times (N_{t} - O_{t})}{N_{t}^2 \times (N_{t} - 1)}$ (@collet). The variance for group 2 is calculated similarly, and the total variance is calculated by the formula $V_{t} = V_{1t} + V_{2t}$. @tbl-logrank2 shows the calculation of the Chi-Square test statistic for the data set. In the table, $V1t$ is the variance of the number of events at time t in group 1, $V2t$ is the variance of the number of events at time t in group 2, $Vt$ is the total variance of the number of events across both groups at time t, $sum_{Vt}$ is the cumulative sum of the total variance across all times, and $X2$ is the Chi-Square test statistic.

```{r}
#| label: tbl-logrank2
#| tbl-cap: "Calculation of Variance"

log_rank$V1t <- (log_rank$N1t * log_rank$N2t * log_rank$Ot * 
                   (log_rank$Nt - log_rank$Ot)) / 
  (log_rank$Nt^2 * (log_rank$Nt - 1))

log_rank$V2t <- (log_rank$N1t * log_rank$N2t * log_rank$Ot * 
                   (log_rank$Nt - log_rank$Ot)) / 
  (log_rank$Nt^2 * (log_rank$Nt - 1))

round(log_rank, 2) %>% 
  gt() %>% 
  tab_header("Calculation of Variance") %>% 
  cols_label(Time = "Time", 
             N1t = "Number at Risk in Group 1", 
             N2t = "Number at Risk in Group 2", 
             Nt = "Total Number at Risk", 
             O1t = "Observed Events in Group 1", 
             O2t = "Observed Events in Group 2", 
             Ot = "Total Observed Events", 
             E1t = "Expected Events in Group 1", 
             E2t = "Expected Events in Group 2", 
             V1t = "Variance of Number of Events in Group 1", 
             V2t = "Variance of Number of Events in Group 2")
```

For group 1 at time 1, we would calculate $E_{1t} = \frac{5 \times 1}{10} = 0.50$ and for group 2 at time 1, $E_{2t} = \frac{5 \times 1}{10} = 0.50$. We would repeat for each group at each time and then sum the values to get $\sum_{t=1}^t{E_{i}}$ for each group. For group 1, $\sum_{t=1}^t{E_{1}} = 0.50 + 0.56 + 0 + 0 + 0.67 + 0.80 + 0.75 + 0 = 3.27$ and for group 2, $\sum_{t=0}^t{E_{2}} = 0.50 + 0.44 + 0 + 0 + 0.33 + 0.20 + 0.25 + 0 = 1.73$. The sum of the observed events for group 1 is 2 and for group 2 is 3. For group 1 at time 1, we would calculate $V_1t = \frac{5 \times 5 \times 1 \times (10-1)}{10^2 \times (10-1)} = 0.25$ @tbl-chi-square shows the rest of the variances calculated for both groups for the `surv2` data set.

```{r}
#| label: tbl-chi-square
#| tbl-cap: "Calculation of Chi Square Statistic"

sum_O1t <- cumsum(log_rank$O1t)
sum_O2t <- cumsum(log_rank$O2t)
sum_E1t <- cumsum(log_rank$E1t)
sum_E2t <- cumsum(log_rank$E2t)
sum_V1t <- cumsum(log_rank$V1t)
sum_V2t <- cumsum(log_rank$V2t)

# Create table with each of the last values 
sum_stats <- data.frame(sum_O1t, sum_O2t, 
                        sum_E1t, sum_E2t, sum_V1t, sum_V2t)
# Get the last row of summary statistics
sum_stats <- sum_stats[nrow(sum_stats), ]
sum_stats$X2_group1 <- ((sum_stats$sum_O1t - sum_stats$sum_E1t)^2 / 
                          sum_stats$sum_V1t)
sum_stats$X2_group2 <- ((sum_stats$sum_O2t - sum_stats$sum_E2t)^2 / 
                          sum_stats$sum_V2t)

round(sum_stats, 2) %>% 
  gt() %>% 
  tab_header("Calculation of Chi Square Statistic") %>% 
  cols_label(sum_O1t = "Sum of Observed Events in Group 1", 
             sum_O2t = "Sum of Observed Events in Group 2", 
             sum_E1t = "Sum of Expected Events in Group 1", 
             sum_E2t = "Sum of Expected Events in Group 2", 
             sum_V1t = "Sum of Variance of Number of Events in Group 1", 
             sum_V2t = "Sum of Variance of Number of Events in Group 2", 
             X2_group1 = "Chi Square Statistic for Group 1", 
             X2_group2 = "Chi Square Statistic for Group 2")
```

With all of these values, we can calculate the Chi Square test statistic, $X^2i = \frac{(\sum{O_{it}} - \sum{E_{it}})^2}{V_{it}}$. We get $X^2group1 = \frac{(2 - 3.27)^2}{1.07} = 1.51$ and $X^2group1 = \frac{(3 - 1.73)^2}{1.07} = 1.51$. The test statistic can then be compared to a Chi-Square distribution with one k-1 degrees of freedom, with k being the number of groups. So, for this example, there is one degree of freedom. If the p-value is less than 0.05, then we can reject the null hypothesis that there is no difference between the survival estimates of the two groups (@sullivan16).

Clearly, calculating the test statistic is very tedious, even for a data set with 10 observations. The `survdiff` function can be used to run a Log-Rank test in R, making it much easier to run the test. The function takes in the response object created by the `Surv()` function and the grouping variable (@zabor23). It returns a Chi-Square statistic and a p-value. We can see that the p-value is much higher than 0.05, indicating that we do not have evidence to reject the null hypothesis, and thus that there is no difference between the survival estimates of the two groups.

```{r}
# Calculate times for events in this data set
times2 <- Surv(surv2$time, surv2$status)
# Run a log rank test based on group variable
test_stat <- survdiff(times2 ~ group, data = surv2)
test_stat
```

The only adjustment needed for plotting the Kaplan Meier curves for two groups is to include the grouping variable while making the survival object using the `survfit` function (@zabor23). The two survival curves are depicted in @fig-survcurv2.

```{r}
#| label: fig-survcurv2
#| fig-cap: "Kaplan Meier Survival Curve by Group"

# Create survival object using group as a predictor
s2 <- survfit(Surv(time, status) ~ group, data = surv2)

# Plot the two curves
s2 %>% 
  ggsurvfit() +
  labs(
    x = "Days",
    y = "Overall survival probability",
    title = "Kaplan Meier Survival Curve by Group"
  ) + 
  scale_x_continuous(breaks=seq(0,10,by=2)) + 
  scale_y_continuous(breaks=seq(0,1,by=.2)) +
  add_risktable()
```

## Case Study

The next example uses a data set with variables for predicting death in patients with cirrhosis, which is permanent scarring of the liver. The data set is from a clinical trial conducted by the Mayo Clinic from 1974 to 1984. The data set contains 424 primary biliary cirrhosis patients with 20 variables (@fedesoriano). To conduct survival analysis, the `Status` variable needs to be transformed into an indicator variable, labelled `event`, coded as a 1 representing death and 0 representing censored. The `N_Days` variable will be used for the time variable, indicating number of days since the beginning of the trial.

```{r}
cirrhosis <- read_csv("data/cirrhosis.csv", show_col_types = FALSE)
cirrhosis <- cirrhosis %>% mutate(event = if_else(Status == 'D', 1, 0)) 
```

The main interest of this data is to explore the difference in survival time between patients taking the drug of interest, D-penicillamine, and those given a placebo. The variable `Drug` indicates which group the patient was in and will be used as the grouping variable for a Log-Rank Test. Below, we calculate a survival object for this data and then use the `survdiff` function to run a Log-Rank Test.

```{r}
# Calculate times and survival object
times3 <- Surv(cirrhosis$N_Days, cirrhosis$event)
s3 <- survfit(times3 ~ Drug, data = cirrhosis)
# Run a log rank test based on drug group
survdiff(times3 ~ Drug, data = cirrhosis)
```

From the Log-Rank test, we see that the p-value is 0.7, which is mush larger than 0.05. This indicates that the test was not statistically significant at the five percent level. Thus, there was no significant difference between patient outcome between the two treatment groups. For patients with biliary cirrhosis, D-penicillamine is not an effective drug for preventing death.

We can again visualize the difference in survival curves in @fig-casestudy.

```{r}
#| label: fig-casestudy
#| fig-cap: "Kaplan Meier Survival Curve Predicting Death from Cirrhosis"

s3 %>% 
  ggsurvfit() +
  labs(
    x = "Time",
    y = "Overall survival probability",
    title = "Kaplan Meier Survival Curve"
  ) + 
  add_risktable()
```

It is important to note the limitations of this analysis. The Kaplan Meier Curve is not a suitable fit for a complex, real-world data set such as the `cirrhosis` data. It is a nonparametric method of analysis, which means that it assumes no form of distribution and no additional factors influencing the outcome of interest. To create a Kaplan Meier curve and run the Log-Rank test, we ignored all other factors included int the data. In the next section, we will discuss other methods that can be used for survival analysis that can account for other factors.

## References
